---
title: "Convergence loop"
author: Mauricio Garnier-Villarreal
bibliography:
  - refs.bib
---

```{r pkgld, include=FALSE}
library(blavaan, quietly=TRUE)
library(lavaan, quietly=TRUE)
```


### Introduction

In many cases you would need to run the BSEM models multiple times until it has converged. This can take a while and you might want to have *R* do it for you. This tutorial shows how to use a ```while``` loop to increase the number of burnin samples until the model converges, so you can let it run without having to adjust it every time

### Convergence loop

You will start by writing the model syntax as always. Then instead of running the ```blavaan``` functions as usual, we will run then inside the ```while``` loop as follows. 

Before the loop starts you need to define a starting ```BURN <- 0``` number of iterations, and a convergence value higher than the desired such as ```rhat <- 20```. 

Then the loop will be set sto stop when the convergence criteria (```rhat```) is lower than a desired value, like $\hat{R} < 1.05$, we specify this with ```while(rhat > 1.05)```, meaning the the loop will continue as long as rhat is higher than 1.05. 

```{r, eval=T, include=FALSE, cache=TRUE}
HS.model <- ' visual  =~ x1 + x2 + x3
              textual =~ x4 + x5 + x6
              speed   =~ x7 + x8 + x9 '

BURN <- 0
rhat <- 20
while(rhat > 1.05) {
  
  BURN <- BURN + 1000 ### increase burn in by 1000 iterations every time
  
  fit <- bcfa(HS.model, std.lv=T, 
              data=HolzingerSwineford1939, 
              n.chains = 3, burnin = BURN,
              sample=1000)
  rhat <- max(blavInspect(fit, "psrf"), na.rm=T)
  print(paste0("Rhat=",rhat))  
}

```

```{r, eval=F}
HS.model <- ' visual  =~ x1 + x2 + x3
              textual =~ x4 + x5 + x6
              speed   =~ x7 + x8 + x9 '

BURN <- 0
rhat <- 20
while(rhat > 1.05) {
  
  BURN <- BURN + 1000 ### increase burn in by 1000 iterations every time
  
  fit <- bcfa(HS.model, std.lv=T, 
              data=HolzingerSwineford1939, 
              n.chains = 3, burnin = BURN,
              sample=1000)
  rhat <- max(blavInspect(fit, "psrf"), na.rm=T)
  print(paste0("Rhat=",rhat))  
}

```


Thn inside the loop we will increase the number of ```BURN``` iterations by 1000 in this example. And after estimating the model, we will evaluate the convergence by getting the highest estimated $\hat{R}$, and printing it in the screen so you will see how far is the model from converging. 

```{r, eval=T}
print(paste0("Rhat=",rhat))  
```

Note that we are only increasing the number burnin iterations, and keeping the number of saved samples the same (1000 in this case). If you want you can increase or decrease the number of saved iterations according to your case. 

And you can visualize the convergence with the trace plots

```{r}
plot(fit, pars = 1:9, plot.type = "trace")
```


### Convergence criteria 

In this example we use $\hat{R} < 1.05$ as the convergence criteria. We recommend you to use this or $\hat{R} < 1.01$ as the convergence criteria, but not higher. 

As $\hat{R}$ approximates 1, we can argue that the model has converged as the estimates have achieve stability between and within chains [@gelman_bayesian_2014]


### References