---
title: "Plot Functionality"
bibliography:
  - refs.bib
---

```{r pld, include=FALSE}
library(blavaan, quietly=TRUE)
library(lavaan, quietly=TRUE)
library(ggplot2, quietly=TRUE)
library(bayesplot, quietly=TRUE)
```

```{r modfit, include=FALSE, cache=TRUE}
HS.model <- ' visual  =~ x1 + x2 + x3
              textual =~ x4 + x5 + x6
              speed   =~ x7 + x8 + x9 '
     
fit <- bcfa(HS.model, data=HolzingerSwineford1939, seed=959)
```

A `plot()` method exists for *blavaan* objects, with this method making use of the *bayesplot* package [@bayesplot]. We provide details here about how this functionality works. We will use a 3-factor model for demonstration:

```{r moddisp, eval=FALSE}
HS.model <- ' visual  =~ x1 + x2 + x3
              textual =~ x4 + x5 + x6
              speed   =~ x7 + x8 + x9 '
     
fit <- bcfa(HS.model, data=HolzingerSwineford1939)
```

### Basics
Because many *blavaan* models will have many parameters, users generally need to specify which parameters they wish to plot. This is accomplished by supplying numbers to the `pars` argument, where the numbers correspond to the order of parameters from the `coef()` command (the numbers also appear in the `free` column of the parameter table). Users must also specify the type of plot that they desire via the `plot.type` argument. So, for example, a trace plot of the first four model parameters looks like

```{r p1}
plot(fit, pars = 1:4, plot.type = "trace")
```

Many other plot types are available, coming from the *bayesplot* package. In general, for *bayesplot* functions that begin with `mcmc_`, the corresponding `plot.type` is the remainder of the function name without the leading `mcmc_`. Examples of many of these plots can be found in [this bayesplot vignette](https://cran.r-project.org/web/packages/bayesplot/vignettes/plotting-mcmc-draws.html).


### Customization
Users may wish to customize some aspects of the resulting plots. For this, the `plot()` function will output a *ggplot* object. This makes it possible to modify the plot as if it were any other *ggplot* object, which allows for many possibilities. One starting point for exploring *ggplot2* is [here](https://ggplot2.tidyverse.org/).

```{r p2}
p <- plot(fit, pars = 1:4, plot.type = "trace", showplot = FALSE)

p + facet_text(size=15) + legend_none()
```

Alternatively, users may wish to create a plot that is entirely different from what is available via `plot()`. This can be facilitated by extracting the posterior samples or the Stan model, via `blavInspect()`:

```{r eval=FALSE}
## list of draws
## (one list entry per chain):
draws <- blavInspect(fit, "mcmc")

## convert the list to a matrix
## (each row is a sample,
##  each column is a parameter)
draws <- do.call("rbind", draws)

## Stan (or JAGS) model
modobj <- blavInspect(fit, "mcobj")
```


### References
