---
title: "Prior Predictive Checks"
author: Mauricio Garnier-Villarreal
bibliography:
  - refs.bib
---

```{r pkgld, include=FALSE}
library(blavaan, quietly=TRUE)
library(lavaan, quietly=TRUE)
library(bayesplot, quietly=TRUE)
load("prior_pred_out.RData")
```


### Introduction

In Bayesian models we need to specify priors for the model parameters. Priors are the distribution that we *think* the parameters would follow, even before having data. These can represent **high** or **low** uncertainty, so that a *diffuse* prior indicates that we don not know a lot about how the parameter would behave, while an *informative* prior means that we are quite certain about the expected distribution.

### Prior Predictive Checks

Prior predictive checks (PPC) generate data according to the prior in order to asses whether a prior is appropriate [@Gabry_2019_vis]. A posterior predictive check generates replicated data according to the posterior predictive distribution. In contrast, the prior predictive check generates data according to the prior predictive distribution $y^{sim} ∼ p(y)$.

The prior predictive distribution is just like the posterior predictive distribution with no observed data, so that a prior predictive check is nothing more than the limiting case of a posterior predictive check with no data.

This is easy to carry out mechanically by simulating parameters $θ^{sim}∼p(\theta)$ according to the priors, then simulating data $y^{sim}∼p(y∣ \theta^{sim})$  according to the sampling distribution given the simulated parameters. The result is a simulation from the joint distribution, $(y^{sim},θ^{sim})∼p(y,\theta)$ and thus $y^{sim}∼p(y)$ is a simulation from the prior predictive distribution.

In *blavaan* we can get PPC when we use the argument ```prisamp=TRUE``` , this will tell blavaan to ignore the data and buil distributions only from the priors. Here we will start by adjusting the priors, instead of using the default priors.

#### Weakly informative priors

We will show an example with the @holswi39 data, first a case with weakly informative priors. Here we are stpecifying that the observeded variable intercepts will have a prior of $N(3, 2)$, the factor loadings will have a prior of $N(0.4, 2)$, and the residual standard deviation with a prior of $\Gamma(1,1)$. 

```{r, eval=FALSE}
priors <- dpriors(nu="normal(3,2)",
                  lambda="normal(0.4, 2)",
                  theta="gamma(1,1)[sd]")
```

Then we estimate the BSEM model with the respective priors on the ```dp``` argument, and with ```prisamp=TRUE```, so getting PPC instead of posterior distributions.

```{r, eval=F}
HS.model <- ' visual  =~ x1 + x2 + x3
              textual =~ x4 + x5 + x6
              speed   =~ x7 + x8 + x9 '

fit_wi <- bcfa(HS.model, data=HolzingerSwineford1939, 
            std.lv=TRUE, meanstructure=T, test = "none",
            dp=priors, prisamp = T)
```

You might get some warning messages from either divergent and/or failed convergence. For this we would ignore these messages as it is likely to have issues for the evaluations of prior predictions. 

We now have a **blavaan** object with prior predictive distributions, so we can use any of the package functions to describe them, and see if the parameters are within expected ranges. For example we can get the PPC density distributions for the first 9 parameters (factor loadings in this case). The basic ```plot()``` method calls the functions from the ```bayesplot``` package [@bayesplot] and with the ```plot.type = "dens"``` argument we can plot the density distributions

```{r, eval=T}
plot(fit_wi, pars=1:9, plot.type = "dens")
```

You can also pick which parameters to plot, like the factor correlations by chossing parameters ```19:21``` in this case

```{r, eval=T}
plot(fit_wi, pars=19:21, plot.type = "dens")
```

From the factor loadings distributions we see that the first loading on each factor is bounded by 0, this is due to a modeling identification constraint in blavaan, and with a maximum values aroun 6. And for all other distributions they range between -6 to 6 or -4 to 4, and for all priors the most likely value is around 0. This would be described as weakly informative priors as it allows them to range between begative and positive values but without allowing crazy high/low values 

Note that what would be a *realistic* range would dependen on the parameter, model specification, and data. So, consider these priors in function of those characterictics. 

For the factor correlations we kept the deafult diffuse priors, so these allowed from very high and low correlation, but the prior distributions are not flat across all possible correlation values. 

#### Default priors

In this next example, we will estimate the PPC with the package default priors, that would consider diffuse priors. You can see the blavaan default priors with the function ```dpriors()```

```{r, eval=T}
dpriors()
```

Then we estimate the BSEM model and ignore the ```dp``` argument letting run with the default priors, and with ```prisamp=TRUE```, so getting PPC instead of posterior distributions.

```{r, eval=F}
fit_df <- bcfa(HS.model, data=HolzingerSwineford1939, 
            std.lv=TRUE, meanstructure=T, test = "none", 
            prisamp = T)
```

Then we can plot the density distributions and compare them. We see that with the default diffuse priors, the model allows up to very high values such as -30 to 30

```{r, eval=T}
plot(fit_df, pars=1:9, plot.type = "dens")
```

This way we can see that the more diffuse priors allows a higher range of values. It is up to the researcher to decide which range of priors better present their expectations. 

It is important to note that these PPC allows to see the expected distributions based on the priors, but these might not be the same as the priors used in the estimation process, as the priors interact with the model specification and constraints (such as the o bound constraint for the first factor loading) [@merkle2023opaque]

### References